% Final Report: AI-Powered Intrusion Detection System (IDS) using Machine Learning
% LLNCS format - Springer
\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,calc}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\algnewcommand\algorithmicinput{\textbf{Input:}}
\algnewcommand\Input{\item[\algorithmicinput]}
\algnewcommand\algorithmicoutput{\textbf{Output:}}
\algnewcommand\Output{\item[\algorithmicoutput]}

\begin{document}

\title{An AI-Powered Intrusion Detection System: Real-Time Detection of Network Attacks using Machine Learning on the CIC-IDS2017 Dataset}

\titlerunning{ML-Based IDS: Real-Time Detection of Network Attacks}

\author{Le Anh Tuan\inst{1}\thanks{Student ID: 252805005} \and
        To Hoang Dao\inst{1}\thanks{Student ID: 252805401} \and
        Nguyen Tu Thanh Duy\inst{1}\thanks{Student ID: 252805701}}

\institute{Ton Duc Thang University (TDTU), Ho Chi Minh City, Vietnam}

\maketitle

\begin{abstract}
Modern enterprise networks face an ever-expanding attack surface, with adversaries employing techniques ranging from volumetric Distributed Denial-of-Service (DDoS) floods to stealthy credential brute-force and web-layer injection attacks. Signature-based Intrusion Detection Systems (IDSs) remain effective against known threats but fail to generalise to novel and polymorphic attacks. This paper presents an AI-powered IDS that applies supervised machine learning to classify network flows across 13 traffic classes using the CIC-IDS2017 benchmark. We design an end-to-end pipeline comprising stratified sampling, a proportional SMOTE oversampling strategy based on a square-root target formula, and comparative training of four classifiers: LightGBM, XGBoost, Random Forest, and Decision Tree. Experimental results on a held-out test set of 46,838 flows show that \textbf{LightGBM with Proportional SMOTE} achieves the best Macro-$F_1$ of \textbf{0.9241} and Balanced Accuracy of \textbf{0.9253}, with six of thirteen classes reaching perfect $F_1 = 1.00$. All tree-based models sustain $> 30{,}000$ flows per second on commodity CPU hardware, meeting the latency budget for inline deployment. We identify the remaining failure modes---chiefly the confusion between Web Attack sub-types and the degraded reliability of classes with over 80\% synthetic training samples---and discuss targeted mitigations including payload-level features and threshold tuning.

\keywords{Intrusion Detection System \and Machine Learning \and CIC-IDS2017 \and LightGBM \and XGBoost \and Random Forest \and SMOTE \and Class Imbalance \and Network Security}
\end{abstract}

\input{1/1}

\input{2/2}

\input{3/3}

\input{4/4}

\input{5/5}
\input{6/6}


%=======================================================================
\clearpage
\begin{thebibliography}{8}

\bibitem{ref_cicids2017}
Sharafaldin, I., Habibi Lashkari, A., Ghorbani, A.A.: Toward Generating a New Intrusion Detection Dataset and Intrusion Traffic Characterization. In: Proc.\ 4th Int.\ Conf.\ on Information Systems Security and Privacy (ICISSP), pp.\ 108--116 (2018)

\bibitem{ref_kdd99}
Tavallaee, M., Bagheri, E., Lu, W., Ghorbani, A.A.: A Detailed Analysis of the KDD CUP 99 Data Set. In: Proc.\ IEEE Symp.\ on Computational Intelligence for Security and Defense Applications (CISDA), pp.\ 1--6 (2009)

\bibitem{ref_rf}
Breiman, L.: Random Forests. Machine Learning \textbf{45}(1), 5--32 (2001)

\bibitem{ref_xgboost}
Chen, T., Guestrin, C.: XGBoost: A Scalable Tree Boosting System. In: Proc.\ 22nd ACM SIGKDD, pp.\ 785--794 (2016)

\bibitem{ref_smote}
Chawla, N.V., Bowyer, K.W., Hall, L.O., Kegelmeyer, W.P.: SMOTE: Synthetic Minority Over-Sampling Technique. JAIR \textbf{16}, 321--357 (2002)

\bibitem{ref_dl_ids}
Vinayakumar, R., Alazab, M., Soman, K.P., Poornachandran, P., Al-Nemrat, A., Venkatraman, S.: Deep Learning Approach for Intelligent Intrusion Detection System. IEEE Access \textbf{7}, 41525--41550 (2019)

\bibitem{ref_anomaly_survey}
Buczak, A.L., Guven, E.: A Survey of Data Mining and Machine Learning Methods for Cyber Security Intrusion Detection. IEEE Commun.\ Surv.\ Tutor.\ \textbf{18}(2), 1153--1176 (2016)

\bibitem{ref_tensorflow}
Abadi, M., et al.: TensorFlow: A System for Large-Scale Machine Learning. In: Proc.\ 12th USENIX OSDI, pp.\ 265--283 (2016)

\bibitem{ref_lightgbm}
Ke, G., Meng, Q., Finley, T., Wang, T., Chen, W., Ma, W., Ye, Q., Liu, T.Y.: LightGBM: A Highly Efficient Gradient Boosting Decision Tree. In: Advances in Neural Information Processing Systems (NeurIPS), vol.\ 30, pp.\ 3146--3154 (2017)

\bibitem{ref_cicflowmeter}
Lashkari, A.H., Gil, G.D., Mamun, M.S.I., Ghorbani, A.A.: Characterization of Tor Traffic using Time based Features. In: Proc.\ 3rd Int.\ Conf.\ on Information Systems Security and Privacy (ICISSP), pp.\ 253--262 (2017). \url{https://www.unb.ca/cic/research/applications.html\#CICFlowMeter}

\bibitem{ref_snort}
Roesch, M.: Snort---Lightweight Intrusion Detection for Networks. In: Proc.\ 13th USENIX LISA Conference, pp.\ 229--238 (1999)

\bibitem{ref_adasyn}
He, H., Bai, Y., Garcia, E.A., Li, S.: ADASYN: Adaptive Synthetic Sampling Approach for Imbalanced Learning. In: Proc.\ IEEE Int.\ Joint Conf.\ Neural Networks, pp.\ 1322--1328 (2008)

\bibitem{ref_focal}
Lin, T.Y., Goyal, P., Girshick, R., He, K., Doll\'{a}r, P.: Focal Loss for Dense Object Detection. In: Proc.\ IEEE ICCV, pp.\ 2980--2988 (2017)

\bibitem{ref_prauc}
Saito, T., Rehmsmeier, M.: The Precision--Recall Plot is More Informative than the ROC Plot when Evaluating Binary Classifiers on Imbalanced Datasets. PLoS ONE \textbf{10}(3), e0118432 (2015)

\end{thebibliography}

\end{document}
