\subsection{The CIC-IDS2017 Dataset}
\label{sec:bg-dataset}

\subsubsection*{Overview and Collection Methodology.}
The \textbf{CIC-IDS2017} dataset, released by the Canadian Institute for Cybersecurity, was built specifically to address the well-known deficiencies of older benchmarks (KDD'99, NSL-KDD) such as redundant records, obsolete attack vectors, and unrealistic traffic patterns~\cite{ref_cicids2017,ref_kdd99}. It was collected over \emph{five working days} (Monday through Friday, 3--7 July 2017) on a dedicated testbed comprising a victim network of 10 hosts running Windows, Linux, and macOS, and an attacker network of 4 machines launching scripted campaigns. Benign background traffic was generated via the \emph{B-Profile} framework, which statistically models user behaviour across HTTP, HTTPS, FTP, SSH, and email protocols.

The raw PCAP captures were processed with the \textsc{CICFlowMeter} toolkit~\cite{ref_cicflowmeter}, which emits one row per bidirectional flow and labels each flow with the ground-truth class derived from the attack schedule. The resulting corpus contains approximately \textbf{2.83 million labelled flows}, distributed across one benign class and fourteen attack classes (commonly collapsed to seven aggregate classes for modelling).

\subsubsection*{Feature Semantics.}
CIC-IDS2017 provides \textbf{78 numeric features} per flow (plus the label). For the purpose of this report we group them into four semantically meaningful families:

\begin{description}[leftmargin=1.2em,itemsep=2pt,topsep=2pt]
    \item[Flow-level descriptors.] \texttt{Flow Duration}, \texttt{Flow Bytes/s}, \texttt{Flow Packets/s}, \texttt{Total Fwd/Bwd Packets}, \texttt{Total Length of Fwd/Bwd Packets}. These capture the gross size and rate of the exchange---large values indicate volumetric traffic such as DDoS or file transfer; near-zero values hint at reconnaissance probes.
    \item[Packet-size statistics.] \texttt{Fwd/Bwd Packet Length Min, Max, Mean, Std}, \texttt{Packet Length Variance}, \texttt{Min/Max Packet Length}. Distributions of packet sizes discriminate, for example, short-packet SYN floods from long-packet bulk transfers.
    \item[Inter-arrival times (IAT).] \texttt{Flow IAT Mean/Std/Max/Min}, \texttt{Fwd IAT Total/Mean/Std}, \texttt{Bwd IAT Total/Mean/Std}, \texttt{Active/Idle Mean/Std}. Timing features are the strongest signal for periodic beaconing (C2), brute-force loops, and low-and-slow scans.
    \item[TCP flag and header counts.] \texttt{SYN/ACK/FIN/RST/PSH/URG Flag Count}, \texttt{Fwd/Bwd PSH/URG Flags}, \texttt{Fwd/Bwd Header Length}. Flag distributions are especially informative for detecting half-open connections, scans, and malformed handshakes.
\end{description}

\begin{table}[h]
\centering
\caption{Aggregate class distribution of CIC-IDS2017 (approximate).}
\label{tab:class-dist}
\begin{tabular}{lrr}
\toprule
\textbf{Class} & \textbf{\#Flows} & \textbf{\%} \\
\midrule
BENIGN                     & 2,273,097 & 80.30 \\
DDoS                       &   128,027 &  4.52 \\
PortScan                   &   158,930 &  5.62 \\
DoS Hulk                   &   231,073 &  8.16 \\
DoS GoldenEye              &    10,293 &  0.36 \\
DoS slowloris / Slowhttptest &  11,420 &  0.40 \\
FTP-Patator / SSH-Patator   &    13,835 &  0.49 \\
Web Attack (Brute/XSS/SQLi) &     2,180 &  0.08 \\
Bot                         &     1,966 &  0.07 \\
Infiltration                &        36 &  0.0013 \\
Heartbleed                  &        11 &  0.0004 \\
\midrule
\textbf{Total}              & \textbf{2{,}830{,}868} & \textbf{100} \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection*{Dataset Statistics and Challenges.}
Table~\ref{tab:class-dist} summarizes the class distribution. Two properties dominate the modelling discussion. First, the dataset is \textbf{severely imbalanced}: the benign class alone accounts for $\approx 80\%$ of all flows, whereas classes such as \texttt{Heartbleed} and \texttt{Infiltration} contain fewer than a hundred examples each. Naive training therefore produces classifiers that default to \texttt{BENIGN} and achieve deceptively high overall accuracy while failing on the minority attack classes that matter most. Second, the feature set contains a small number of columns with undefined values (\texttt{NaN}, $\pm\infty$) produced by edge cases in the flow calculator (e.g., division by zero for single-packet flows), and a handful of zero-variance columns that must be removed. These issues motivate the preprocessing pipeline described next.
