\subsection{Discussion of Hard Cases}
\label{sec:exp-scenarios}

\subsubsection*{Web Attack Confusion: XSS vs.\ Brute Force.}
The systematic confusion between \texttt{Web Attack -- XSS} and \texttt{Web Attack -- Brute Force} is the most operationally significant failure mode in our pipeline. Both attack types target the HTTP layer and produce flows of similar duration, packet count, and IAT distribution. At the flow level, the only distinguishing signal lies in subtle payload-derived statistics (e.g., entropy of response sizes, URI length distribution) that CICFlowMeter does not capture. All four models in the consensus test misclassified the XSS sample as Brute Force, confirming that this confusion is not model-specific but feature-space-inherent. A practical mitigation would be to merge the two Web Attack sub-types into a single \texttt{Web Attack} class for alert generation, accepting a coarser label in exchange for reliable detection.

\subsubsection*{Rare\_Attack: Heterogeneous Grouping.}
The \texttt{Rare\_Attack} group merges three attack types with qualitatively different network signatures: \texttt{Infiltration} (low-volume exfiltration beaconing), \texttt{SQL Injection} (HTTP POST-heavy flows), and \texttt{Heartbleed} (specific TLS record-layer anomaly). Merging them into a single label forces the classifier to learn a single decision boundary for three distinct distributions, inevitably increasing confusion with both \texttt{BENIGN} and Web Attack classes. The consensus test showed that two models correctly predicted \texttt{Rare\_Attack} while the other two predicted \texttt{Web Attack -- XSS} and \texttt{BENIGN}, reflecting this heterogeneity. With more data (or dedicated payload-level features), these classes would warrant separate labels and individual model evaluation.

\subsubsection*{Bot Detection.}
The \texttt{Bot} class (1,948 samples in the original corpus) achieves $F_1 = 0.86$ with LightGBM after SMOTE. Bot traffic in CIC-IDS2017 is generated by a PCAP replay of the Ares botnet, which produces characteristic HTTP-based C2 beaconing with periodic inter-arrival times. The moderately high recall (0.80) suggests the model captures the temporal regularity, but the approximately 80\% synthetic fraction introduces some label noise that depresses precision. Deploying the IDS with a lower decision threshold for the Bot class (at the cost of a slightly elevated FPR on BENIGN) would increase recall toward the operationally desirable value of $> 0.95$.
